A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}

### 2022-01-26 tex/esop22/syntax.tex
[ esop22 ] Reorder Section 6, try to smooth narrative
[ fix ] references
[ edit ] Some minor tidy ups in Section 4

We take the insights of the previous section and use them to build a
generic framework for posemiring-annotated substructural systems in
Agda. We will first show \emph{descriptions} of systems, which are
comprised of rules that have premises combined using the bunched
combinators. We then show how to construct the Agda data type of
intrinsically well scoped, typed, and resourced terms for any given
system of our framework. We use the prototypical system from
\figref{fig:comb-lr} as a running example. \secref{sec:other-syntaxes}
presents further examples that our framework can express.

@@

\paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}

A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
Each \AgdaRecord{Rule} comprises a tree of \AgdaDatatype{Premises} and
a type of conclusion. We assume that there is a
$\AgdaBound{Ty} : \AgdaPrimitiveType{Set}$ of types for the system in
scope.

The \AgdaDatatype{Premise} data type describes premises of rules,
using the bunched combinators from the previous section. A single
premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
constructor.  This allows binding of additional variables
\AgdaBound{$\Delta$} (with specified types and usage annotations) and
the specification of a conclusion type \AgdaBound{A} for this premise.
The remaining constructors are descriptions for the first-order
bunched connectives, and will be interpreted directly as such, below.

@@

\paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}

@@

\paragraph{Terms of a \AgdaRecord{System}.}

@@

With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
the rest are simply interpretations of premises into bunched combinators.
The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
they have been lifted from predicates on usage vectors to predicates on
contexts, with the type component of the context shared throughout.
Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
already polymorphic (not relying on anything specific about usage vectors), so
do not need a \AgdaFunction{$^c$} variant.

@@

The interpretation of a \AgdaRecord{Rule} checks that the rule targets
the desired type and then interprets the rule's premises \AgdaBound{ps}.
Notice that the interpretation of the premises is independent of the conclusion
of the rule, which accounts for the difference in type between
\AgdaFunction{$\llbracket$\_$\rrbracket$p} and
\AgdaFunction{$\llbracket$\_$\rrbracket$r}.

@@

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is where we
use sized types to record that subterms are smaller than the
surrounding term.

@@

This definition uses \AgdaFunction{$\dotto$}, which, analogously to
\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
space.
We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
case two (the context and the type).
Informally,
\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
stands for
\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
\AgdaSpace{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
where \AgdaBound{T} is a type family with $n$ many indices.

Terms defined like this are still quite difficult to write, mainly because of
frequently changing usage contexts and the need for proofs that they all match
up.
We will see a way of automating these proofs in \cref{sec:usage-elaborator}.

>>>>

We take the insights of the previous section and use them to build a
generic framework for posemiring-annotated substructural systems in
Agda. We will first show \emph{descriptions} of systems, which are
comprised of rules that have premises combined using the bunched
combinators. We then show how to construct the Agda data type of
intrinsically well scoped, typed, and resourced terms for any given
system of our framework. We use the prototypical system from
\figref{fig:lr-comb} as a running example. \secref{sec:other-syntaxes}
presents further examples that our framework can express.

@@

% \paragraph{\AgdaDatatype{Premises}, \AgdaRecord{Rule}s, and \AgdaRecord{System}s.}

A type \AgdaRecord{System} is made up of multiple \AgdaRecord{Rule}s.
Each \AgdaRecord{Rule} comprises a \AgdaDatatype{Premises} and
a conclusion type. We assume that there is a
$\AgdaBound{Ty} : \AgdaPrimitiveType{Set}$ of types for the system in
scope.

The \AgdaDatatype{Premise} data type describes premises of rules,
using the bunched combinators from \cref{sec:bunched-rules}. A single
premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
constructor.  This allows binding of additional variables
\AgdaBound{$\Delta$} (with specified types and usage annotations) and
the specification of a conclusion type \AgdaBound{A} for this premise.
The remaining constructors are descriptions for the
bunched connectives. %, and will be interpreted directly as such, below.

@@

% \paragraph{\figref{fig:lr-comb} as a \AgdaRecord{System}.}

@@

\subsection{Terms of a \AgdaRecord{System}.}\label{sec:terms}

@@

The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
are the interpretation into bunched combinators.
The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
they have been lifted from predicates on usage vectors to predicates on
contexts, with the type component of the context shared throughout.
Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
already polymorphic (not relying on anything specific about usage vectors), so
do not need a \AgdaFunction{$^c$} variant.

@@

The interpretation of a \AgdaRecord{Rule} checks that the rule targets
the desired type and then interprets the rule's premises
\AgdaBound{ps}.  Notice that the interpretation of the premises is
independent of the conclusion of the rule, which accounts for the use
of \AgdaFunction{OpenType} in
\AgdaFunction{$\llbracket$\_$\rrbracket$p} versus
\AgdaFunction{OpenFam} in \AgdaFunction{$\llbracket$\_$\rrbracket$r}.

@@

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is a use of
Agda's sized types to record that subterms are smaller than the
surrounding term for the termination checker.

@@

This definition uses \AgdaFunction{$\dotto$}, which, analogously to
\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
space.
We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
case two (the context and the type).
The notation
\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
stands for
\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
\AgdaSpace{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
where \AgdaBound{T} is a type family with $n$ many indices.

Terms defined like this are difficult to write by hand, due to the
need for proofs that the usage contexts are handled correctly for the rules used.
We will see a way of automating these proofs in \cref{sec:usage-elaborator}.
